A Closer Look at Time-Preserving Paths on Evolving Graphs

### 2017-10-05 paper.tex
add more content

A time-preserving path on evolving graph makes implicit assumptions about the type of edges in which a path traverse. For example, some measures count only static edges,
while some measures take account of both static edges and causal edges. In addition, the length of a time-preserving path can be either spatial or temporal.

We study the impact of different time-preserving paths on node centrality. We compare evolving graph centrality with the aggregated static graph case to gain insight about the advantage of evolving graph model. We construct an evolving coauthor network from
a collection of research papers for the experiments.
\end{abstract}

@@

In the 1985 science-fiction film ``Back to the Future'', Marty McFly (played by Michael J. Fox) traveled back in time and accidentally changed the history.
However, we cannot travel back in time or even send a message to the past.
An evolving graph in general is a time dependent graph $G(t) = (V(t), E(t))$.
To traverse an evolving graph, one has to consider time. In particular, a path on an evolving graph needs to be time-preserving, meaning a node can not pass a massage (through an edge) to a node at a earlier time.

% time-respecting path
We represent an evolving graph as a time-ordered sequences of graphs, similar to the work of Tang and coworkers \cite{tang10, tang102, tang09, nicosia13} and Grindrod, Higham and coworkers \cite{grindrod11, grindrod13}. The idea is to divide the system's time span into temporal slices and regards it as a sequence of static graphs $\{G^{[t]}\}$, one for each layer. A node on a slice $G^{[t]}$ is represented by a pair $(v, t)$, where $v$ is a node of $G^{[t]}$ and $t$ is the time stamp of the slice. A time-respecting path of length $n$ is a sequence of these node pairs $\langle (v_1, t_1), (v_2, t_2), \ldots ,(v_m, t_m)\rangle$, where $t_1 \le t_2 \le \cdots \le t_n$. Note if $t_i = t_j$, then $\langle (v_i, t_i), (v_j, t_j) \rangle$ is an edge at a temporal slice,  otherwise it is an edge linking nodes from different temporal slices. 
We call edges of the first kind \emph{static edges} and of the second kind
\emph{causal edges}.
For example, in Figure \ref{fig:eg_shortest_path}
$\langle (1, t_1), (2, t_1) \rangle$ is a static edge at time stamp $t_1$ and $\langle (3, t_2), (3, t_3) \rangle$ is a causal edge between time stamp $t_2$ and time stamp $t_3$.
The difference between static edges and causal edges were introduced in \cite{chen16}.

@@

It is natural to define the length of a path on a static graph as the number of nodes traveled through the path. For the weighted graph case, the length of a path is the total weight of all the edges in the path. We observe that static edges link nodes in space while causal edges link nodes in time. Hence we would like to distinguish temporal distance from spatial distance on an evolving graph: the length of a static edge is equal to the edge weight (or $1$ in the unweighted graph case); the length of a causal edge is the time distance between the two nodes. In the simplest case, the time-respecting path
$\langle (1, t_1) ,(2, t_1) , (2, t_3), (3, t_3)\rangle$ in Figure \ref{fig:eg_shortest_path} has length $4$ in total because two static edges $\langle (1, t_1) (2, t_1) \rangle$ and $\langle (2, t_3), (3, t_3) \rangle$ have length one, whereas $\langle (2, t_1) (2, t_3) \rangle$ traverse $2$ time stamps so has length $2$.

@@

% our contribution this paper
Borgatti noted in \cite{borgatti05} that the manner of traffic flow entails a new way to think about centrality. In particularly, the importance of a node in a network can not be determined without reference to how traffic flows through the network. A time-preserving path on evolving graphs can be described by static edges, causal edges, or both.
Grindrod et al.'s definition of dynamic walk \cite{grindrod11} only count static edges. 
Tang et al \cite{tang10s} measures a temporal shortest path in time unites (or causal edges) only. If a node can reached by another node in $k$ time stamps then the distance between the two nodes is $k$. Chen and Zhang \cite{chen16} considered both static edges and causal edges in their definition of temporal path. 
To develop a centrality algorithm on evolving graphs, one has to consider the manner of temporal traffic flow.

@@

The betweenness centrality of a node $i$ in a static graph is defined as
$$
C_i = \sum_{j \in V}\sum_{k \in V, k \ne j} \frac{\rho_{jk}(i)}{\rho_{jk}},
$$
where $\rho_{jk}$ is the number of shortest paths from node $j$ to node $k$, while $\rho_{jk}(i)$ is the number of shortest paths from node $j$ to node $k$ that pass through node $i$.

% \subsection{Eigenvector Centrality}
% \label{sec:eigenv-centr}

@@

We review a few well-known measures of evolving graph centrality.

\subsection{Temporal Katz Centrality}
\label{sec:temp-katz-centr}

In \cite{grindrod11}, Grindrod et. al. defines a \emph{dynamic walk} of length $m$ from
node $v_1$ to node $v_{w+1}$  as a sequence of edges 
$(v_1, v_2), (v_2, v_3), \ldots (v_w, v_{w+1})$ and a non-decreasing sequence of times
$t_{r_1} \leq t_{r_2} \leq \cdots \leq t_{r_w}$ such that the adjacency matrix at time stamp
$r_m$, i.e., $A_{i_m, i_{m+1}}^{[r_m]} \ne 0$. Note in this definition, the authors only take account of the static edges and the length of a dynamic walk is determined by the number of static edges.

@@

@@

\subsection{Temporal PageRank}
\label{sec:temporal-pagerank}

@@

We can represent an evolving graph $G_n$ by a static graph $G$.

Here we discuss different types of temporal flow on the evolving graphs.
\begin{itemize}
\item Only static edges.
\item Only causal edges.
\item Both causal edges and static edges.
\item Reverse temporal flows.
\item Active nodes.
\end{itemize}

\subsection{Only Static Edges}
\label{sec:only-static-edges}

\subsection{Only Causal Edges}
\label{sec:only-causal-edges}

In \cite{tang10s}, a temporal shortest path is measured in time units only, i.e., if node
$v_i$ can reach node $v_j$ in $k$ time steps then the distance between node
$v_i$ and $v_j$ is $k$. If we define temporal path in this way (only take accounts of the time stamps), we have a new centrality definition. 
We can use temporal breadth first
search \cite{chen16} to discover all the temporal paths and then only take account of the difference between the time stamp of the first node and the time stamp of the last node.

In matrix form, the nonzero entries of $A^{[t_i]} A^{[t_{i+1}]}$ counts all the temporal paths of length $1$, and 
$A^{[t_i]}\cdots A^{[t_j]}$, where $i < j$ counts all the temporal paths of length $j -i$. 
For example, $A^{[1]}A^{[4]}$, $A^{[1]}A^{[1]}A^{[4]}$ and $A^{[1]}A^{[2]}A^{[3]}A^{[4]}$ are both walks of length 3.
In the original Katz centrality, the `attenuation' factor $\alpha$ is imposed on the
number of walks, i.e., longer walks has smaller effect to overall rating. In this case, the final result is the summation of all products of the form
\[
\alpha^k A^{[i]} \cdots A^{[i+k]}
\]
Note this can not be written in \eqref{eq:katz}. The eigenvector centrality vector $v$ is the limit of
\[
v = \lim_{k \rightarrow \inf}\big[ \frac{1}{k} (A + \lambda_1^{-1} A^2 + \lambda_1^{-2}A^3 + \cdots + \lambda_1^{-k}A^{k+1})\big].
\]
Here we can replace $A^2$ or $A^3$ by $A^{[1]}A^{[2]}$ for the evolving graph case.
Since PageRank is a eigenvalue problem of a weighted matrix $G$ (instead of $A$),
 we can also express PageRank in this form. There has some work on writing evolving graph Pagerank as a sequence of PageRank matrix product.

\subsection{Both Static Edges and Causal Edges}
\label{sec:both-static-edges}

Another definition of temporal path was introduced by \cite{chen16}. In their definition,
both static edges and causal edges are considered. We can written the evolving graph centrality as a big static block matrix and Katz centrality of this big matrix. This gives you the importance of each node at different time stamps. We can `average' these ratings at different time stamp to get the final results.

Let's see a simple example. For the evolving graph shown below

@@

In this section we describe some illustrative

\subsection{Synthetic Data}
\label{sec:synthetic-data}

@@

A second network is

author names, affiliation, journal, and year of publication.

We are interested in finding out how good evolving graph centrality algorithms
comparing with static graph ones. 
Also how the way to

stability, reasonable solutions that can be applied in practice.

@@

Evolving graph model is

We conduct the experiments on a list of papers with associated information, including author names, affiliation, journal, and year of publication

Evaluating and comparing the performance of centrality algorithms is hard as there is no formal definition for the ``qualities'' of a person.

>>>>

A time-preserving path on evolving graphs makes implicit assumptions about the type of edges in which a path traverse. For example, some papers count only static edges,
while some take account of both static edges and causal edges. In addition, the length of a time-preserving path can be either spatial or temporal.

We study the impact of different time-preserving paths on node centrality by varying the edge weights of static and causal edges.
We also compare evolving graph centrality with the aggregated static graph case to gain insight about the advantage of evolving graph model.
We construct an evolving coauthor network from a collection of research papers for the experiments.
\end{abstract}

@@

In the 1985 science-fiction film ``Back to the Future'', Marty McFly (played by Michael J. Fox) travelled back in time and accidentally changed the history.
However, we cannot travel back in time or even send a message to the past.
An evolving graph in general is a time dependent graph $G(t) = (V(t), E(t))$.
To traverse an evolving graph, one has to consider time. In particular, a path on an evolving graph needs to be time-preserving, meaning a node can not pass a massage (through an edge) to a node at a earlier time.

% time-respecting path
We represent an evolving graph as a time-ordered sequences of graphs, similar to the work of Tang and coworkers \cite{tang10, tang102, tang09, nicosia13} and Grindrod, Higham and coworkers \cite{grindrod11, grindrod13}. The idea is to divide the system's time span into temporal slices and regards it as a sequence of static graphs $G^{[t]}$, one for each layer. A node on a slice $G^{[t]}$ is represented by a pair $(v, t)$, where $v$ is a node of
$G^{[t]}$ and $t$ is the time stamp of the slice. A time-respecting path of length $n$ is a sequence of these node pairs $\langle (v_1, t_1), (v_2, t_2), \ldots ,(v_m, t_m)\rangle$, where $t_1 \le t_2 \le \cdots \le t_n$. Note if $t_i = t_j$, then $\langle (v_i, t_i), (v_j, t_j) \rangle$ is an edge at a temporal slice,  otherwise it is an edge linking nodes from different temporal slices.
We call edges of the first kind \emph{static edges} and of the second kind
\emph{causal edges}.
For example, in Figure \ref{fig:eg_shortest_path}
$\langle (1, t_1), (2, t_1) \rangle$ is a static edge at time stamp $t_1$ and $\langle (3, t_2), (3, t_3) \rangle$ is a causal edge between time stamp $t_2$ and time stamp $t_3$.
The difference between static edges and causal edges were introduced in \cite{chen16}.

@@

It is natural to define the length of a path on a static graph as the number of nodes travelled through the path. For the weighted graph case, the length of a path is the total weight of all the edges in the path. We observe that static edges link nodes in space while causal edges link nodes in time. Hence we would like to distinguish temporal distance from spatial distance on an evolving graph: the length of a static edge is equal to the edge weight (or $1$ in the unweighted graph case); the length of a causal edge is the time distance between the two nodes. In the simplest case, the time-respecting path
$\langle (1, t_1) ,(2, t_1) , (2, t_3), (3, t_3)\rangle$ in Figure \ref{fig:eg_shortest_path} has length $4$ in total because two static edges $\langle (1, t_1) (2, t_1) \rangle$ and $\langle (2, t_3), (3, t_3) \rangle$ have length one, whereas $\langle (2, t_1) (2, t_3) \rangle$ traverse $2$ time stamps so has length $2$.

@@

% our contribution this paper
Borgatti noted in \cite{borgatti05} that the manner of traffic flow entails a new way to think about centrality. In particularly, the importance of a node in a network can not be determined without reference to how traffic flows through the network.
To develop a centrality algorithm on evolving graphs, one has to consider the manner of temporal traffic flow.
A time-preserving path on evolving graphs can be described by static edges, causal edges, or both.
Grindrod et al.'s definition of dynamic walk \cite{grindrod11} only count static edges.
Tang et al. \cite{tang10s} measures a temporal shortest path in time unites (or causal edges) only. If a node can reached by another node in $k$ time stamps then the distance between the two nodes is $k$. Chen and Zhang \cite{chen16} considered both static edges and causal edges in their definition of temporal path.

@@

The \emph{betweenness centrality} of a node $i$ in a static graph is defined as
$$
C_i = \sum_{j \in V}\sum_{k \in V, k \ne j} \frac{\rho_{jk}(i)}{\rho_{jk}},
$$
where $\rho_{jk}$ is the number of shortest paths from node $j$ to node $k$, while $\rho_{jk}(i)$ is the number of shortest paths from node $j$ to node $k$ that pass through node $i$

% The \emph{communicability betweenness centraltiy} \cite{estrada09} for a node $i$ is defined as
% $$
% C_i = C_N\sum_p \sum_q \frac{\exp(A)_{p,q} - \exp(A - E(i))_{p,q}}{\exp(A)_{p,q}},
% $$
% where $C_N = \frac{1}{(N-1)^2 - (N-1)}$ is a normalizing factor and $E(i)$
% has nonzeros only in row and column $i$, and in such row and column has $-1$ wherever $A$ has $1$. Similarly, the \emph{resolvent betweenness centrality} for node $i$ is defined as
% $$
% C_i = C_N\sum_p \sum_q \frac{(I - \alpha A)^{-1}_{p,q} - (I - \alpha(A - E(i)))^{-1}_{p,q}}{(I - \alpha A)^{-1}_{p,q}}.
% $$
% \subsection{Eigenvector Centrality}
% \label{sec:eigenv-centr}

@@

In \cite{tang10s}, a temporal shortest path is measured in time units only, i.e., if node
$v_i$ can reach node $v_j$ in $k$ time steps then the distance between node
$v_i$ and $v_j$ is $k$. We can apply temporal breadth first
search (BFS) \cite{chen16} to discover the temporal paths and then only take account of the temporal difference between the first node $(v_1, t_1)$ and the last node $(v_n, t_n)$, i.e., $t_n - t_1$.

If we use this definition of temporal path for the temporal Katz centrality introduced by
\cite{grindrod11}, we have a new centrality measure.
In the matrix form, the nonzero entries of $A^{[t_i]} A^{[t_{i+1}]}$ counts all the temporal paths of length $1$, and
$A^{[t_i]}\cdots A^{[t_j]}$, where $i < j$ counts all the temporal paths of length $j -i$.
For example, $A^{[1]}A^{[4]}$, $A^{[1]}A^{[1]}A^{[4]}$ and $A^{[1]}A^{[2]}A^{[3]}A^{[4]}$ are all walks of length 3.
In the original Katz centrality, the `attenuation' factor $\alpha$ is imposed on the
number of walks, i.e., longer walks has smaller effect to the overall rating.
Hence the final result is the summation of all products of the form
\[
\alpha^k A^{[i]} \cdots A^{[i+k]}
\]
Note this can not be written in \eqref{eq:katz}, as discussed in Section \ref{sec:temp-katz-centr}.

Another definition of temporal path was introduced by \cite{chen16}. Here
both static edges and causal edges are considered. We can represent an evolving graph with a block adjacency matrix. Centraliy on this block matrix gives the importance of each node at all time stamps. We can average these rating at different time stamps to
get the final rating.

We review a few well-known measures of evolving graph centrality.
(Design a simple example and discuss it in each case below.)
\subsection{Temporal Katz Centrality}
\label{sec:temp-katz-centr}

In \cite{grindrod11}, Grindrod et al. defines a \emph{dynamic walk} of length $m$ from
node $v_1$ to node $v_{w+1}$  as a sequence of edges
$(v_1, v_2), (v_2, v_3), \ldots (v_w, v_{w+1})$ and a non-decreasing sequence of times
$t_{r_1} \leq t_{r_2} \leq \cdots \leq t_{r_w}$ such that the adjacency matrix at time stamp
$r_m$, i.e., $A_{i_m, i_{m+1}}^{[r_m]} \ne 0$. Note in this definition, the authors only take account of the static edges and the length of a dynamic walk is determined by the number of static edges.

@@

Recall that betweenness centrality of a node in a static graph is defined by counting the fraction of shortest paths that traverse it.
We could extend
betweenness centrality on evolving graphs by replacing shortest paths with temporal shortest paths \cite{nicosia13}.
\begin{equation}
  \label{eq:bet1}
C_i^{betweenness} = \sum_{j \in V}\sum_{k \in V, k \ne j}\frac{\alpha_{jk}(i)}{\alpha_{jk}},
\end{equation}
where $\alpha_{jk}$ represents the number of temporal shortest paths from node $j$ to node $k$ and $\alpha_{jk}(i)$ represents the number of temporal shortest paths from node $j$ to node $k$ that pass through the node $i$.

One may want to take into account the length of time a node trains a message. The temporal betweenness centrality of the node $i$ at time stamp $t_m$ is defined as
\begin{equation}
  \label{eq:bet2}
  C_i^{betweenness}(t_m) = \sum_{j\ne i}\sum_{k\ne j, k\ne i}\frac{U_{j,k,t_m}(i)}{\alpha_{j,k}},
\end{equation}
where as before $\alpha_{j,k}$ represents the number of temporal shortest paths from $j$ to $k$ and $U_{j,k,t_m}(i)$ represents the number of temporal shortest paths from $j$ to $k$ in which node $i$ is traversed from
the path in the time stamp $t_m$ or a previous time stamp $t_{m-1}$. Then
average temporal betweenness of node $i$ is then defined as
$$
  C_i^{betweenness} = \frac{1}{M}\sum_m C_i^{betweenness}(t_m).
$$
Note \eqref{eq:bet1} only takes account of static edges while \eqref{eq:bet2} takes account of both static edges and causal edges.

@@

Let $d_{ij}$ is the length of temporal shortest path from $i$ and $j$ in an evolving graph. Then
the closeness centrality of a node $i$ is defined
$$
C_i^{closeness} = \frac{N-1}{\sum_j d_{ij}}.
$$
Note depends on how we define the length of temporal shortest path, the closeness centrality can takes account of only static edges, only causal edges, or both.

@@

In \cite{chen16}, we derive a block adjacency matrix representation.
Let $E'$ be the set of causal edges and $\hat E$ be the set of static edges then
an evolving graph can be represented as
$$
\bm M_n =
\begin{pmatrix}
A^{[t_1]} & M^{[t_1, t_2]} & \ldots & M^{[t_1, t_n]} \\
0         & A^{[t_2]} & \ldots & M^{[t_2, t_n]} \\
          & \ldots    &        &     \\
0         & 0         & \ldots & A^{[t_n]}
\end{pmatrix},
$$
where $M^{[t_i, t_j]}$ is the matrix whose rows are labeled by $V^{[t_i]}$ and columns are labeled by $V^{[t_j]}$, and whose entries are
$$
  M_{uv}^{[t_i, t_j]} =
  \begin{cases}
    1 & \mbox{if} (u, v) \in E' \\
    0 & \mbox{otherwise}.
  \end{cases}
$$
The adjacency matrix blocks $A^{[t]}$ encode the static edge set $\hat E$, whereas the off-diagonal blocks $M^{[t_i, t_j]}$ encode the causal edge set $E'$.

We can apply static graph centrality algorithms to $M_n$ and then take the average of ratings at each each time stamp. For example, we can construct the Google matrix
$G_n$ from $M_n$ and the ranking of the nodes can be found by solving
$$
 r= G_n r.
$$

For Figue \ref{fig:eg_shortest_path}, we have

\begin{align*}
V & = \{(1,\!t_1), (2,\!t_1), (1,\!t_2), (3,\!t_2), (2,\!t_3), (3,\!t_3)\},\\
\tilde E &= \{((1,\!t_1), (2,\!t_1)), ((1,\!t_2), (3,\!t_2)), ((2,\!t_3), (3,\!t_3))\},\\
E'  &= \{((1,\!t_1), (1,\!t_2)), ((2,\!t_2), (2,\!t_3)), ((3,\!t_2), (3,\!t_3))\}.
\end{align*}

and corresponding block adjacency matrix is

\[
\bm M_3 = \begin{pmatrix}
0 & 1 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 0
\end{pmatrix}
\]

The static graph representation of Figure \ref{fig:eg_shortest_path} is shown in Figure \ref{fig:static}.

\begin{figure}[h]
 \begin{center}
   \begin{tikzpicture}[scale=.36, line width =1.4pt]
  \node[circle,fill=green1, minimum size=0.1cm, inner sep = 0pt] (n7) at (-5,7)
{\scriptsize $(2,t_1)$};
  \node[circle,draw=red1, minimum size=0.2cm, inner sep = 0pt] (n8) at (-10,5)
{\scriptsize $(3,t_1)$};
  \node[circle,fill=green1, minimum size=0.1cm, inner sep = 0pt] (n10) at (-8,10) {\scriptsize $(1,t_1)$};

  \node[circle,draw=red1, minimum size=0.2cm, inner sep = 0pt] (n6) at (3,7)
{\scriptsize $(2,t_2)$};
  \node[circle,fill=green1, minimum size=0.2cm,  inner sep = 0pt] (n4) at (-2,5)
{\scriptsize $(3,t_2)$};
  \node[circle,fill=green1, minimum size=0.2cm,  inner sep = 0pt] (n1) at (0,10)
{\scriptsize $(1,t_2)$};

   \node[circle,fill=green1, minimum size=0.2cm,  inner sep = 0pt] (n11) at (11,7)
{\scriptsize $(2,t_3)$};
  \node[circle,fill=green1, minimum size=0.2cm,  inner sep = 0pt] (n12) at (6,5)
{\scriptsize $(3,t_3)$};
  \node[circle,draw=red1, minimum size=0.2cm,  inner sep = 0pt] (n14) at (8,10)
{\scriptsize $(1,t_3)$};

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=80, in=40] (n11);
  \draw[dotted,->](n10) to[out=10, in=180] (n1);
   \draw[dotted,->](n4) to[out=-30,in=-150] (n12);
    \end{tikzpicture}
\end{center}
\caption{A static graph corresponding to the evolving graph example of
Figure~\ref{fig:eg_shortest_path}. The green nodes are active nodes while the
red nodes are inactive nodes.
The black lines are edges in the static edge set $\tilde E$ and are encoded
algebraically in the diagonal blocks $A^{[t]}$ of the adjacency matrix $\bm M_3$.
The dotted lines are edges in the causal edge set $E'$ and are encoded algebraically in
the off-diagonal blocks $M^{[t_i, t_j]}$.
The graph containing all the edges and temporal nodes has adjacency matrix $\bm M_3$.}
\label{fig:static}
\end{figure}

@@

In this section, we vary the edge
weights in both static and causal edges to study the impact of time-preserving paths on evolving graph centrality.
We also compare evolving graph centrality with the aggregated static graph case to gain insight about the advantage of evolving graph model.

@@

(add results here.)

A second network is constructed by linking leading authors in different papers whose
ideas are similar. The resulting network represent the connectivity of ideas in all the research papers in the database.

(add results here.)

@@
